---
title: "Data-Driven Approach To Improving Hospital OU Operations"
author: "Uyen Tran & Tam Nguyen"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
setwd("~/Desktop/Data Driven Decision Making/Final Project")
df <- read.csv("OUData.csv") 
```

```{r}
summary(df)
```

```{r}
str(df)
```
# Transform Variable Types
```{r}
# Variable Transformation
df$Gender <- as.factor(df$Gender)
df$PrimaryInsuranceCategory <- as.factor(df$PrimaryInsuranceCategory)
df$InitPatientClassAndFirstPostOUClass <- as.factor(df$InitPatientClassAndFirstPostOUClass)
df$Flipped <- as.factor(df$Flipped)
df$DRG01 <- as.factor(df$DRG01)
df$BloodPressureUpper <- as.numeric(df$BloodPressureUpper)
df$BloodPressureLower <- as.numeric(df$BloodPressureLower)
df$BloodPressureDiff <- as.numeric(df$BloodPressureDiff)
df$Pulse <- as.numeric(df$Pulse)
df$PulseOximetry <- as.numeric(df$PulseOximetry)
df$Respirations <- as.numeric(df$Respirations)
df$Temperature <- as.numeric(df$Temperature)
```


```{r}
# Checking for duplicates
sum(duplicated(df))
```

```{r}
# Check for missing values
library(kableExtra)
table.missing<-data.frame(miss.val=sapply(df, function(x)
  sum(length(which(is.na(x))))))
table.missing%>%kable(caption = "Number of missing values")%>%kableExtra::kable_styling(bootstrap_options = "striped")
```


## Checking distribution of numerical variables with missing values

```{r}
#  Histogram of "PulseOximetry" after scaling
hst <- hist(df$PulseOximetry,
     main = "Histogram of PulseOximetry",
     xlab = "PulseOximetry",
     ylab = "Frequency",
     col = "lightblue",
     border = "blue")
hst
```

```{r}
#  Histogram of "Respirations" after scaling
hst1 <- hist(df$Respirations,
     main = "Histogram of Respirations",
     xlab = "Respirations",
     ylab = "Frequency",
     col = "lightblue",
     border = "blue")
hst1
```


```{r}
#  Histogram of "BloodPressureUpper" after scaling
hst3 <- hist(df$BloodPressureUpper,
     main = "Histogram of BloodPressureUpper",
     xlab = "BloodPressureUpper",
     ylab = "Frequency",
     col = "lightblue",
     border = "blue")
hst3
```

```{r}
#  Histogram of "Pulse" 
hst4 <- hist(df$PulseOximetry,
     main = "Histogram of Pulse",
     xlab = "Pulse",
     ylab = "Frequency",
     col = "lightblue",
     border = "blue")
hst4
```

```{r}
#  Histogram of "Temperature" 
hst4 <- hist(df$Temperature,
     main = "Histogram of Temperature",
     xlab = "Temperature",
     ylab = "Frequency",
     col = "lightblue",
     border = "blue")
hst4
```

### Replace NA with median since the distribution of these variables are skewed

```{r}
library(dplyr)
df <- df %>%
  mutate_at(vars(PulseOximetry, Respirations, BloodPressureDiff, BloodPressureUpper, Pulse, Temperature), ~ifelse(is.na(.), median(., na.rm = TRUE), .))

print(df)
```

```{r}
# Checking missing values in all column
colSums(is.na(df)) > 0
```

### Remove unuseful columns
```{r}
# Remove unuseful variables
df <- subset(df, select = -c(ObservationRecordKey, InitPatientClassAndFirstPostOUClass))
df
```


## Data Visualization


```{r}
library(ggplot2)
# Histogram of Age, colored by Flipped
ggplot(df, aes(x = Age, fill = Flipped)) +
  geom_histogram(position = "identity", alpha = 0.9, bins = 20, color = "black") +
  labs(title = "Distribution of Age by Flipped Status",
       x = "Age",
       y = "Frequency") +
  scale_fill_manual(values = c("white", "navy"), labels = c("Not Flipped", "Flipped")) +
  theme_minimal()
```

**DRG code and Flipping Rate**
```{r}
# Calculate the proportion of patients flipped to inpatient status by DRG code
proportion_data <- df %>%
  group_by(DRG01) %>%
  summarise(Proportion_Inpatient = mean(Flipped == 1, na.rm = TRUE)) 

# Sort the data by count in descending order to highlight the highest
proportion_data <- proportion_data %>%
  arrange(desc(Proportion_Inpatient))

# Reorder DRG01 factor levels based on Proportion_Inpatient
proportion_data$DRG01 <- factor(proportion_data$DRG01, levels = proportion_data$DRG01)

# Create the bar chart
ggplot(proportion_data, aes(x = DRG01, y = Proportion_Inpatient)) +
  geom_bar(stat = "identity", fill = "navy", color = "white") +
  labs(title = "Proportion of Patients Flipped to Inpatient Status by DRG Code",
       x = "DRG Code",
       y = "Proportion of Inpatients") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )
```
**Box plots**
```{r}
library(gridExtra)

# Vital Sign and Flipping Rate
p1 <- ggplot(df, aes(x = Flipped, y = PulseOximetry)) +
  geom_boxplot(fill = "navy", color = "black") +
  labs(title = "Relationship between Pulse Oximetry and Flipping Rate",
       x = "Inpatient Status",
       y = "Pulse Oximetry") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )

# Length of Stay and Flipping Rate
p2 <- ggplot(df, aes(x = Flipped, y = OU_LOS_hrs)) +
  geom_boxplot(fill = "navy", color = "black") +
  labs(title = "Relationship between Length of Stay and Flipping Rate",
       x = "Inpatient Status",
       y = "Length of Stay") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )

# Blood Pressure Difference and Flipping Rate
p3 <- ggplot(df, aes(x = Flipped, y = BloodPressureDiff)) +
  geom_boxplot(fill = "navy", color = "black") +
  labs(title = "Relationship between Blood Pressure Difference and Flipping Rate",
       x = "Inpatient Status",
       y = "Blood Pressure Difference") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )

# Temperature and Flipping Rate
p4 <- ggplot(df, aes(x = Flipped, y = Temperature)) +
  geom_boxplot(fill = "navy", color = "black") +
  labs(title = "Relationship between Temperature and Flipping Rate",
       x = "Inpatient Status",
       y = "Temperature") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )

# Arrange the plots 
grid.arrange(p1, p2, p3, p4, 
             ncol = 2)

```


```{r}
# Bar chart of count of flipped patient by primary insurance company
flipped_counts <- df %>% 
  group_by(PrimaryInsuranceCategory, Flipped) %>%
  summarise(Count = n()) %>%
  ungroup()

ggplot(flipped_counts, aes(x=PrimaryInsuranceCategory, y = Count, fill = Flipped)) + 
  geom_bar(stat = "identity", position = position_dodge(), color = "black") +
  scale_fill_manual(values = c('white', 'navy'), 
                    labels = c('Not Flipped', 'Flipped'),
                    name = "Flipped Status") +
  labs(x = "Primary Insurance Category", y = "Count", title = "Flipped Status Counts by Primary Insurance Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )
```
```{r}
# Convert Flipped variable to factor to ensure it's treated as categorical
df$Flipped <- factor(df$Flipped, levels = c(0, 1), labels = c("Not Flipped", "Flipped"))

# Calculate percentage of flipped patients within each primary insurance category
flipped_percentages <- df %>%
  group_by(PrimaryInsuranceCategory, Flipped) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100) 

# Plot bar chart with percentages
ggplot(flipped_percentages, aes(x = PrimaryInsuranceCategory, y = Percentage, fill = Flipped)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 2), "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 3, 
            color = "black") +
  scale_fill_manual(values = c('white', 'navy'), 
                    labels = c('Not Flipped', 'Flipped'),
                    name = "Flipped Status") +
  labs(x = "Primary Insurance Category", 
       y = "Percentage of Flipped Patients", 
       title = "Percentage of Flipped Patients by Primary Insurance Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )
```



```{r}
# Scatter plot for Pulse Oximetry vs Length of Stay
ggplot(df, aes(x = OU_LOS_hrs, y = Age)) +
  geom_point(aes(color = as.factor(Flipped))) +
  scale_color_manual(values = c("grey", "navy")) +
  labs(x = "Length of Stay in OU (hours)", y = "Age", 
       color = "Flipped Status", title = "Age vs Length of Stay") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )
```
```{r}
# Calculate the count of flipped status by gender
gender_flipped <- df %>%
  group_by(Gender) %>%
  count(Flipped) %>%
  mutate(Percentage = n / sum(n))

# Generate the stacked bar chart with percentages
ggplot(gender_flipped, aes(x = Gender, y = Percentage, fill = as.factor(Flipped))) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c('white', 'navy'), 
                    labels = c('Not Flipped', 'Flipped'),
                    name = "Flipped Status") +
  labs(x = "Gender", y = "Percentage", title = "Percentage of Flipped Status by Gender") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )
```


**Correlation Matrix of numerical value**
```{r}
library(corrplot)

# Select numerical variables
numerical_vars <- df[, sapply(df, is.numeric)]

# Compute correlation matrix
correlation_matrix <- cor(numerical_vars)

# Plot heatmap of the correlation matrix
corrplot(correlation_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45)
```
## Proportion of the dependent variable "FLipped"
```{r}
# frequency table
tbl <- table(df$Flipped)
round(tbl / nrow(df),2)

# bar chart of frequency for manufacturer
barplot(tbl,
        xlab = "1 = Patient flipped from Observation to Inpatient, 0 = Patient stay in Observation",
        ylab = "frequency",
        main = "Proportion of non-flipped and flipped patients",
        col = "navy")
```
Both categories are proportional, hence there is no need for oversampling.

# Data Partition

```{r}
library(caret)
set.seed(123)
df$Flipped <- ifelse(df$Flipped == "Flipped", 1,0)
# Generate random indexs for training and test sets
train.index <- sample(nrow(df), 0.7*nrow(df))
test.index <- setdiff(seq_len(nrow(df)), train.index)

# Split the data into training (70%) and testing (30%) sets
train.df <- df[train.index,]
test.df <- df[test.index,]

# 
summary(train.df)
summary(test.df)
```

# Logistic Regression

```{r}
# Full model
full.model <- glm(Flipped ~ Age + Gender + PrimaryInsuranceCategory + OU_LOS_hrs + DRG01 + BloodPressureUpper + BloodPressureLower + BloodPressureDiff + Pulse + PulseOximetry + Respirations + Temperature, data = train.df, family = "binomial")

# Stepwise Regression Model
step.model <- step(full.model, direction = "both")
summary(step.model)
```

Using Stepwise elimination method, the best model that achieve the lowest AIC is the model that includes 3 independent variables: Age, PrimaryInsuranceCategory and OU_LOS_hrs.

## Evaluating Predictive Performance of the model

```{r}
pred <- predict(step.model, test.df, type = "response")

# Create a cutoff figure to select the best cutoff value
accT = c()
for (cut in seq(0,1,0.1)){
  cm <- confusionMatrix(factor(ifelse(pred>cut,1,0)), factor(test.df$Flipped))
  accT = c(accT, cm$overall[1])
}
```

```{r}
plot(accT ~ seq(0,1,0.1),main="Figure: Accuracy and Error Rate at Different Cutoffs", xlab = "Cutoff Value", ylab = "", type = "l", ylim = c(0, 1))
lines(1-accT ~ seq(0,1,0.1), type = "l", lty = 2, col="blue")
legend("bottomright", c("accuracy", "overall error"), lty = c(1, 2), merge = TRUE)
```
According to the plot, 0.5 is the best cutoff value to achieve maximum accuracy. Hence, we will use 0.5 as our cutoff for the predictive model.


```{r}
# Confusion matrix - cutoff 0.5
conf.max <- confusionMatrix(factor(ifelse(pred>0.5,1,0)), factor(test.df$Flipped))
conf.max
```
```{r}
# create ROC 
library(pROC)
ROC.logistic <- roc(test.df$Flipped, pred)
# plot ROC Curve
plot(ROC.logistic, print.thres = c(0.5))

#AUC
auc.logistic = ROC.logistic$auc
auc.logistic
```

# Classification Tree
```{r}
library(rpart)
library(tidyr)
default.tree <- rpart(Flipped ~Age + PrimaryInsuranceCategory + OU_LOS_hrs + 
    DRG01, data = train.df, method = "class")
default.tree.pred <- predict(default.tree,test.df,type = "class")

library(rpart.plot)
rpart.plot(default.tree)

# Confusion Matrix
cfm <- confusionMatrix(default.tree.pred, factor(test.df$Flipped))
cfm

# Plot confusion matrix with custom colors
plot(cfm$table, col = c("navy", "white"), main = "Confusion Matrix", 
     xlab = "Actual", ylab = "Predicted")

```
-If LOS >= 55, then the patient is classified as **Flipped**
-Otherwise, If LOS < 55 and DRG = 780,786,787,789, then the patient is predicted as **Non-Flipped**. If this condition is not satisfied, then move on to consider PrimaryInsuranceCategory. If PrimaryInsuranceCategory = Medicaid State or PrimaryInsuranceCategory = Medicare, the patient is predicted as **Non-Flipped**. If this condition is not satisfied, then move on to consider LOS again. If LOS < 26, the patient is predicted as **Non-Flipped**, else the patient is predicted as **Flipped**


```{r}
# ROC curve
library(ROCR)
predict.tree.prob = predict(default.tree, newdata = test.df, type = "prob")

# plot ROC and compute AUC
library(pROC)
ROC.tree = roc(response = test.df$Flipped, predictor = predict.tree.prob[,2])
plot(ROC.tree, print.thres = c(0.5))
auc = ROC.tree$auc
auc
```




# Random Forest

```{r}
library(randomForest)
rf <- randomForest(Flipped ~ Age + PrimaryInsuranceCategory + OU_LOS_hrs + 
    DRG01, data = train.df, ntree = 500, mtry = 4, nodesize = 5, importance = TRUE)

# Plotting varible importance
varImpPlot(rf, type = 1)


# Confusion Matrix
rf.pred <- predict(rf, test.df)
confusionMatrix(factor(rf.pred), factor(test.df$Flipped))
```

# Naive Bayes
```{r}
library(caret)
library(e1071)

# Running Naive Bayes
naive <- naiveBayes(Flipped ~., data = train.df)
naive

# Predicting probabilities 
pred.naive.prob <-predict(rt.naive, newdata = test.df, type="raw")

# Predicting class membership
pred.naive.class <-predict(rt.naive, newdata = test.df)
```

```{r}
#Confusion matrix
confusionMatrix(factor(pred.naive.class), factor(test.df$Flipped))
```




```{r}
## Export results to Excel
#library(openxlsx)
#write.xlsx(cbind(test.df, predict.tree.prob) , file = "test.xlsx")
```

